Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
more

\begin{abstract}
SMT solvers have recently been extended with finite model finding techniques.
However, these cannot cope with universal quantifiers ranging over infinite
types. This paper introduces a translation that eliminates infinite quantifiers
in axioms specifying a large class of recursive functions, including
terminating, tail-recursive, and productive functions. An empirical evaluation
confirms that the approach works well on benchmarks from various sources. The
translation is implemented as a preprocessor in the solver CVC4.
\end{abstract}

@@

Many SMT solvers can reason about quantified formulas using incomplete
instantiation-based methods \cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These methods work well in the context of proving (i.e., showing
unsatisfiability), but they generally cannot be used to find models (i.e.,
showing satisfiability). Typically, a single universal quantifier in one of the
axioms of a problem is enough to prevent the discovery of models.

@@

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used deep inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers. A simple
example where both approaches fail is $\forall n : {\sortint}.\;\, \lite\bigl(
\con{p} \leq 0,\allowbreak\; \con{p}( n ) \teq 1,\allowbreak\; \con{p}( n ) \teq 2 * \con{p}( n - 1 ) \bigr)$.

In this paper, we advocate a method for translating formulas involving function
definitions (such as the one mentioned in the previous paragraph) 
into ones where the techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.

The recursive functions must meet a precise semantic criterion to be admissible
(Section~\ref{sec:prelim}). This criterion is met by terminating recursive
function definitions, but also by tail-recursive functions and even by
productive corecursive functions. The criterion is not met by inconsistent
definitions such as $\forall n : {\sortint}.\;\, \con{f}(n) \teq \con{f}(n) +
1$, but also by some consistent definitions.

Demand for recursive functions in SMT (new draft includes
define-funs-rec). Usually not interested in showing logical consistency, there
exist other tools that check e.g. termination.
Thus, want SMT solver to check satisfiability the assumption that recursive
functions are consistent (where this needs to be defined precisely).

We define an (un)satisfiability-preserving translation method for a class of
formulas containing recursive function definitions (Section~\ref{sec:encoding})
that produces formulas for which existing model finding techniques are
applicable. The idea

replace a definition $\forall x : \tau.\; \const{f}(x) \teq t$ by $\forall a :
\fargsort{\tau}.\; \const{f}(\farg{\con{f}}(a)) \teq t[\concret(a)/x]$, where
$\fargsort{\tau}$ is an uninterpreted ``abstract'' type and $\farg{\con{f}} :
\fargsort{\tau} \to \tau$ a coercion from the abstract type to the concrete
type. For admissible definitions, this preserves satisfiability and
unsatisfiability, while making it possible to find finite models using known
techniques.

We provide experimental evidence that this translation improves the
effectiveness of SMT solvers \cvc and \ziii for finding counterexamples to
verification conditions (Section~\ref{sec:evaluation}).

@@

>>>>

\begin{abstract}
SMT solvers have recently been extended with finite model finding techniques.
However, these cannot cope with universal quantifiers ranging over infinite
types. This paper introduces a translation that eliminates infinite quantifiers
in axioms specifying a large class of recursive functions, including
terminating, tail-recursive, and productive functions. An empirical evaluation
confirms that the approach works well on benchmarks from two sources. The
translation is implemented as a preprocessor in the solver CVC4.
\end{abstract}

@@

Many solvers based on SMT (satisfiability modulo theories) can reason about
quantified formulas using incomplete instantiation-based methods
\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These methods work well in the context of proving (i.e., showing
unsatisfiability), but they generally cannot be used to find models (i.e.,
showing satisfiability). Typically, a single universal quantifier in one of the
axioms of a problem is enough to prevent the discovery of models.

@@

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used deep inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers. A simple
example where both approaches fail is
$$\forall n : {\sortint}.\;\, \lite\bigl(
\con{p} \leq 0,\allowbreak\; \con{p}( n ) \teq 1,\allowbreak\; \con{p}( n ) \teq 2 * \con{p}( n - 1 ) \bigr)$$
This state of affairs is highly unsatisfactory, given the frequence of
recursive definitions in practice and the impending addition of a dedicated
command, \texttt{define-funs-rec}, to the SMT-LIB standard \cite{smtlib25}.

In this paper, we present a method for translating formulas involving function
definitions into formulas where finite model finding techniques can be applied.
The recursive functions must meet a precise semantic criterion to be admissible
(Section~\ref{sec:prelim}). This criterion is met by terminating recursive
function definitions, but also by tail-recursive functions and even by
productive corecursive functions. It is, however, not met by inconsistent
definitions such as $\forall n : {\sortint}.\;\, \con{f}(n) \teq \con{f}(n) +
1$, and it is possible to construct consistent examples that do not meet it.

We define an (un)satisfiability-preserving translation method for a class of
formulas containing admissible recursive function definitions
(Section~\ref{sec:encoding}) that produces formulas for which existing model
finding techniques are applicable. The main insight is that a
recursive definition $\forall x : \tau.\; \const{f}(x) \teq t$
can be translated to
$\forall a :
\fargsort{\tau}.\; \const{f}(\farg{\con{f}}(a)) \teq t[\concret(a)/x]$, where
$\fargsort{\tau}$ is an uninterpreted ``abstract'' type and $\farg{\con{f}} :
\fargsort{\tau} \to \tau$ a coercion from the abstract type to the concrete
type. The translation preserves satisfiability and, for admissible definitions,
unsatisfiability, and it makes finite model finding possible.

Our empirical evaluation on benchmarks from provides evidence that this translation improves the
effectiveness of SMT solvers \cvc and \ziii for finding counterexamples to
verification conditions (Section~\ref{sec:evaluation}).

@@

Admissibility must be checked separately---for example, using a
syntactic criterion or a termination prover.

Usually not interested in showing logical consistency, there
exist other tools that check e.g. termination.
Thus, want SMT solver to check satisfiability the assumption that recursive
functions are consistent (where this needs to be defined precisely).
